---
title: "Practical Machine Learning Prediction Assignment"
author: "Melvin Yap"
output:
  html_document:
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
---

#Executive Summary

Given the training dataset, the objective of this assignment is to develop a
predictive model for target variable **classe** that represents one of the 5
ways that the barbell lifts are performed by the participants. The model is then
tested on the test dataset to determine its accuracy.

#Data Munging

The training and test datasets are first downloaded and loaded as data frames.

```{r, cache=TRUE, results='hide'}
library(RCurl)
urlTrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlTest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
x <- getURL(urlTrain)
y <- getURL(urlTest)
dfTrain <- as.data.frame(read.csv(textConnection(x)))
dfTest <- as.data.frame(read.csv(textConnection(y)))
```

```{r}
#Take note of the number of columns of the datasets
dim(dfTrain)
dim(dfTest)
```

Columns with near-zero variances are then removed as they are not only
non-informative, but may also affect the accuracy of the model.

```{r, cache=TRUE, results='hide'}
library(caret)
nzv_cols <- nearZeroVar(dfTrain)
if(length(nzv_cols) > 0) dfTrain <- dfTrain[, -nzv_cols]
nzv_cols <- nearZeroVar(dfTest)
if(length(nzv_cols) > 0) dfTest <- dfTest[, -nzv_cols]
```

In addition, columns with irrelevant data, e.g. IDs, participant names and 
timestamps are removed from the dataset as they do not value-add to the 
predictive model. In both training and testing datasets, the first six columns
have been identified to be irrelevant and are removed.

```{r, results='hide'}
dfTrain <- dfTrain[7:length(dfTrain)]
dfTest <- dfTest[7:length(dfTest)]
```

Lastly, columns with *NA* values are removed.

```{r, results='hide'}
trainNAs <- apply(dfTrain, 2, function(x) {sum(is.na(x))})
dfTrain <- dfTrain[,which(trainNAs == 0)]
testNAs <- apply(dfTest, 2, function(x) {sum(is.na(x))})
dfTest <- dfTest[,which(testNAs == 0)]
```

```{r}
#Take note of the trimmed number of columns of the datasets
dim(dfTrain)
dim(dfTest)
```

#Partitioning the Data

The training data is partitioned into both training (70%) and cross-validation 
(30%) datasets. The purpose of partitioning is to train the chosen model and 
validate it against the data it was not specifically fitted to in order to
determine the model's accuracy. 

```{r, cache=TRUE, results='hide'}
set.seed(1234)
idxTrain <- createDataPartition(y=dfTrain$classe, p=0.7, list=F)
partTrain <- dfTrain[idxTrain,]
partValidate <- dfTrain[-idxTrain,]
```

#Model Development

##Selecting the Model

Random Forest is an ensemble of decision trees and is selected for its ability 
in finding a natural balance in biased datasets without much tuning, and yet 
produce fairly strong predictions.

##Fitting the Model

The Random Forest model is first fitted with the partitioned training data,
and 

```{r, cache=TRUE, results='hide'}
library(randomForest)
modTrain <- randomForest(classe~., data=partTrain)
```

##Determining the Out-of-Sample Error and Variable Importance

From the output of the training model *modTrain*, it is observed that the out-of-sample
error is represented by the out-of-bag (OOB) error rate of 0.51%, which is small.
The following plot identifies the variables in order of their importance
towards the predictive function. **Roll-belt** is listed as the most important
predictor.

```{r, cache=TRUE}
varImpPlot(modTrain)
```

#Cross-Validation

The model is used to test against the 30% validation dataset, and the respective
**classe** variables are compared to determine its accuracy.

```{r, cache=TRUE}
predictValidate <- predict(modTrain, partValidate)
conMatrix <- confusionMatrix(partValidate$classe, predictValidate)
conMatrix$overall
```

It is noted that the model has an accuracy of **99.66%** within the 95%
confidence interval.

#Testing the Model

The model is now applied to the test dataset and the following prediction
result is obtained, which fits the expected results of this assignment.

```{r, cache=TRUE}
predictTest <- predict(modTrain, dfTest)
predictTest #answers for submission of 20 files
```
